Research Experience

### 2010-03-19 doc/thesis/thesis.tex
worked on particle filter stuff a bit

\section*{Proposed Approach}
\subsection*{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that neuronal
efficacy, $\epsilon$ is a sutiable variable for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

\subsection*{Introduction to Particle Filters}
Particle filters, also known as Sequential Monte-Carlo methods (SMC),
are essentially an on-line version of the Markov-Chain Monte Carlo (MCMC)
algorithm. Similar to MCMC, Particle Filters assume the Markov Assumption
holds, in essence that 
\begin{equation}
P(X_t | X_{t-1}, X_{t-2}, ... X_{0}) = P(X_t | X_{t-1})$
\end{equation}
for a time evolving parameter, $X_t$. This is actually a very good
assumption in most biological systems, since chemicals driving physiological
parameters rarely have memory. Unlike MCMC algorithms, particle filters
only calculate the posterior distribution:
\begin{equation}
P(X_t | y_t, y_{t-1}, ... y_0)
\end{equation}
for $t$ measurements, $y_t, ... y_0$. This frees the algorith from having
to calculate, or even keep track of the values of $X$ at times other than
$t$, and significantly reducing computation time. This also easily lends
the particle filter to solving stochastic differential equations (SDE's).

...

@@

While the General Linear Model has been extremely effective at
gleaning information about which voxels are "active" and which
are not, its purpose is only to say which regions are "statistically
significant". However, considering that voxels consist of millions
of neurons, the reality is that a
region could be active, but at some other level or time constant
than expected; and be completely missed by the linear model. 
Studies showing activation
in Cadavers of Phantoms highlight the inherant problem with the 
General Linear Model: a problem that mandates multiple
comparison testing with very high thresholds to ensure any 
sort of face validity \citep{drift}.
By moving to models that actually "learn" the 
parameters of the model, model error will be drastically
reduced allowing for much more reasonable statistical thresholds.
\subsection{Particle Filters}

@@

Of course, it is well known that many of the assumptions 
that the GLM makes about BOLD activation do not hold. 
The General Linear Model ignores
the fact that the BOLD response is nonlinear, contains non-gaussian
noise and most importantly that brain networks are 
so called "Small World Networks" \citep{smallworld}, \citep{noise}. 
Merely accounting for
nonlinearities in the time-series of the BOLD response can
result in significantly better estimation as shown in \citep{nonlinearmodels}.
But even more crucial is the recent movements away from mass-
univariate models. Current mass-univariate models 
assume uniform connectivity from input to every voxel; yet
this is obviously not the case. While
violating linearity assumptions or Gaussian assumptions 
requires us to be more conservative, the lack of any method
to switch brain circuits on and off means that any model
we make for brain regions would not be Turing complete;
in effect implying that humans are incapable of 
thought \citep{turing}.

So called Dynamic Causal Modeling is the beginning of 
the next phase of neurology. DCM is the first 
brain study to show any significant connection between 
diffusion tensor imaging and actual function layout in
the human brain \citep{dti_dcm}. By incorporating connections
between regions and a realistic activation model, DCM
corrects two of the largest problems with the GLM. While
there are other techniques that may be useful in the future,
they either lack physiological analogs (Artificial Neural
Networks) or are extremely computationally expensive 
(mutual information).  Given that much of the 
potential of the GLM has been exhausted, and that DCM 
is one of the first well defined methods capable of learning 
complicated brain circuits DCM is crucial to the future
of FMRI and our understanding of the human brain.

>>>>

\section*{Proposed Approach}
\subsection*{Goal}
The ultimate goal of this project is to provide a new set of tools
for analyzing FMRI data. Whereas SPM techniques have been highly 
successful at finding macroscopic regions of activation, linear 
modeling can carry significant bias error due to lack of model
flexibility. While adding parameters can significantly increase
error due to model variance, this effect is mitigated by the fact
that we plan to use a model that is based on first principals. The
purpose of this paper is thus to evaluate the potential of using
a particle filter along with the BOLD model to derive physical 
parameters. In so doing, we hope to be able to show that neuronal
efficacy, $\epsilon$ is a suitable variable for estimating voxel 
activation from a standard FMRI image. We also hope to show that 
estimated posterior distribution of the parameters, derived from
the particle filter, is able to provide an accurate measure of the
confidence interval.

\subsection*{Introduction to Particle Filters}
Particle filters, also known as Sequential Monte Carlo (SMC) methods
are a powerful way of estimating the posterior probability distribution
of a set of parameters. Unlike Markov Chain Monte Carlo estimation, particle
filters extend relatively easily to any finite impulse response (FIR) system.
And unlike Variational Bayes or other gradient descent algorithms, particle
filters can be run online, with relatively lower cost. The idea of the particle
filter is to start with a wide mixture PDF of possible parameter sets, 
and then, as measurements come in, to weight more heavily parameter sets 
that tend to give good estimations of the measurements. The reliance on
an initial mixture PDF obviously introduces bias; however, it is often quite
easy to establish a reasonable range of parameters, especially when the
model being used has a physical meaning. Suppose either a set or stream
of measurements $y_t$ are made at discrete times $t = 0, 1, ... $, where
$y_t$ could be a vector of measurements. Then the goal is to find the set
of parameters $\hat{\theta}$ that give the best estimate, $\hat{y}_t$, of $y_t$.
In a biological case, with a model based on first principals, there is 
presumably some true set of parameters, $\theta$, a true model $f$
and a true observation equation, $g$ that dictate how the stimulus 
produce noisy measurements,
\begin{equation}
\dot{x}(t) = f(t, x(t), u(t), \theta) + \nu_d

y(t) = g(t, x(t), u(t), \theta) + \nu_s
\end{equation}

Where $x(t)$ is a vector of state variables, $\theta$ is a vector of system
constants, $\nu_d$ is drift noise (such as a Weiner process), $\nu_s$ is additive noise, $y(t)$ is
the vector of observations. $\nu_d$ may be driven by unknown changes in 
cerebral blood flow, or blood oxygen content, and is assumed to be zero mean.
Similarly $\nu_s$, stationary
noise could be driven by unknown changes due to the heart beat and is also assumed
to be zero mean. There are also non-biological effects that could drive these
noise terms such as inherent MR noise.

Although not generally necessary for particle filters, we will make a few
assumptions based on the particular type of systems faced in biological 
processes. First, that $\theta$ is time invariant. This assumption is 
based on the idea that most physiological parameters don't change in the 
short run unless they are driven by some other stimulus. The obvious exception
might be something like the heart; however, period between heartbeats vary
quickly enough that accurately accounting for heart beats necessitates a 
new set of model parameters, or direct outside information. In short we
assume no parameters are time varying, because not information exists to
describe any of theme in that way. Luckily particle filters are capable 
of dealing with non-white, non-Gaussian noise, suck unanticipated influence
may be re-factored as noise. Secondly we assume that input cannot directly
influence the output, which in the case of the BOLD signal is a good assumption.
%Third, through some sort of preprocessing, we will assume that $\nu_d$ can be
%decreased or completely removed, since it 
Finally, $x(t)$ will include $\theta$, the unknown model constants, although
obviously that will mean that the vector $\dot{x}$ will always have some members
that are 0. The state space equations are then simplified to:
\begin{equation}
\dot{x}(t) = f(x(t), u(t)) + \nu_d

y(t) = g(x(t)) + \nu_s
\end{equation}

The goal of the particle filter, is then to derive the true value of $x(t)$
for each time, given known inputs $u(t), u(t-1), ... u(0)$  and noise 
measurements produced by the system, $y(t), ... , y(0)$ (every time up to
$t$). To begin with then, the particle filter starts with a prior distribution,
which is used to build a Dirac Mixture Probability Distribution function:

\begin{equation}
P(\hat{x}(0)) = \frac{1}{N}\Sigma_0^{N-1} \delta(x_(0) - \hat{x}_i(0))
\end{equation}

where $\hat{x}(0)$ is an empirical representation of the prior, and $\hat{x}_i(0)$
is the $i^{th}$ particle, a single possible representation of the system at time 0.
Initially, since each $\hat{x}_i$ has been drawn form the prior, and no measurements
have been made, the particles are all equally weighted. When a measurement becomes
available, that information is then incorporated into the posterior by re-weighting
the particles .

\begin{equation}
\dot{y} = -ky
\end{equation}
$\theta_i$ would have an estimate for $k$ \emph{and} $y$. Of course in many cases
the model is a potentially non-linear differential equation with no closed
form solution. Initially every $\theta_i$ will have the same weight, because
there isn't yet anything to base a weighting on. Note that this doesn't mean
that the particles have been evenly distributed, so the distribution is not 
necessarily "flat" just because all the particles are equaly weighted. When a
new measurement is available, the particles are then reweighted based on how
well they predicted the new measurement.

@@

While the General Linear Model has been extremely effective at
gleaning information about which voxels are "active" and which
are not, its purpose is only to say which regions are "statistically
significant". However, considering that voxels consist of millions
of neurons, the reality is that a
region could be active, but at some other level or time constant
than expected; and be completely missed by the linear model. 
Studies showing activation
in Cadavers of Phantoms highlight the inherant problem with the 
General Linear Model: a problem that mandates multiple
comparison testing with very high thresholds to ensure any 
sort of face validity \cite{drift}.
By moving to models that actually "learn" the 
parameters of the model, model error will be drastically
reduced allowing for much more reasonable statistical thresholds.
\subsection{Particle Filters}

@@

Of course, it is well known that many of the assumptions 
that the GLM makes about BOLD activation do not hold. 
The General Linear Model ignores
the fact that the BOLD response is nonlinear, contains non-gaussian
noise and most importantly that brain networks are 
so called "Small World Networks" \cite{smallworld}, \cite{noise}. 
Merely accounting for
nonlinearities in the time-series of the BOLD response can
result in significantly better estimation as shown in \cite{nonlinearmodels}.
But even more crucial is the recent movements away from mass-
univariate models. Current mass-univariate models 
assume uniform connectivity from input to every voxel; yet
this is obviously not the case. While
violating linearity assumptions or Gaussian assumptions 
requires us to be more conservative, the lack of any method
to switch brain circuits on and off means that any model
we make for brain regions would not be Turing complete;
in effect implying that humans are incapable of 
thought \cite{turing}.

So called Dynamic Causal Modeling is the beginning of 
the next phase of neurology. DCM is the first 
brain study to show any significant connection between 
diffusion tensor imaging and actual function layout in
the human brain \cite{dti_dcm}. By incorporating connections
between regions and a realistic activation model, DCM
corrects two of the largest problems with the GLM. While
there are other techniques that may be useful in the future,
they either lack physiological analogs (Artificial Neural
Networks) or are extremely computationally expensive 
(mutual information).  Given that much of the 
potential of the GLM has been exhausted, and that DCM 
is one of the first well defined methods capable of learning 
complicated brain circuits DCM is crucial to the future
of FMRI and our understanding of the human brain.
